Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package

### 2017-11-20 docs/manuscript.tex
manuscript editing (added table)

%% recommended packages
\usepackage{thumbpdf,lmodern,amssymb}

% \usepackage[utf8]{inputenc}
% \usepackage{geometry} 
% \geometry{a4paper}  
% %\usepackage[parfill]{parskip} 
% \usepackage{graphicx}	
% \usepackage{amsmath}
% \usepackage{fullpage}
% \usepackage{setspace} 
% \usepackage{lineno}
% \usepackage[none]{hyphenat}
 \usepackage{listings}
% \usepackage[round]{natbib}

@@

In many different contexts, researchers need to make use of data presented in primary literature. Most notably, this includes meta-analysis, which is becoming increasingly common in many research fields. Meta-analysis uses effect size estimates and their sampling variance, taken from many studies, to understand whether particular effects are common across studies and to explain variation among these effects \citep{Naka2017, Glass1976, Koricheva2013, Borenstein2009}. Meta-analysis therefore relies foremost on data extracted from primary literature and more specifically statistics that have been reported in the text or tables of research papers. These summary statistics are also, however, frequently presented in figures and so need to be manually extracted using digitising programs. Although there are several existing tools to perform tasks like this (e.g. DataThief (REF), GraphClick (REF), WebPlotDigitizer (REF))
, these tools are not designed this specific purpose (i.e. a meta-analysis). 
%% need to spell out clear points
%% 1 - dont manipulate data/allow different plot types
Firstly, they typically only provide the user with calibrated x,y coordinates from imported figures, and do not differentiate between common plot types that are use to present data. This means that a large amount of downstream data manipulation is subsequently required. For example, data are frequently presented in mean and error plots (Figure \ref{fig:plot_type}A), for which the user wants a mean and error estimate for each group presented in the figure. With existing programs, x,y coordinates of means and errors are returned, to which the user must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to a standard deviation, depending on the type of error presented.
%% 2- adding metadata
Secondly, digitising programs do not easily allow the integration of metadata at the time of data extraction, such as  grouping points or adding variable names, group names and sample sizes.
%% 3 - not in statistical software
Additionally, existing stand-alone software does not allow easy import into commonly used statistical software (such as \proglang{R}), or try to optimize the research pipelines to facilitate editing and reproducibility. These are major issues because extracting from figures can be an incredibly time-consuming process. Furthermore, although meta-analysis is an important tool in consolidating the data from many studies, relying on transparency of those studies, many of the processes involved in data extraction are opaque and difficult to reproduce, making extending studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and go a long way to resolving the reproducibility crises we are seeing in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, sandve_ten_2013}.

@@

\begin{table}[t!]
\centering
\begin{tabular}{lcccc}
\hline
Function                  & metaDigitise & GraphClick & DataThief  & WebPlotDigitizer\\
\hline
Groups                    & \checkmark   & \checkmark & $\times$   & \checkmark  \\
Scatterplots              & \checkmark   & \checkmark & \checkmark & \checkmark  \\
Mean and error plots      & \checkmark   &            & \checkmark & \\
Boxplots                  & \checkmark   &            & $\times$    & \\
Histograms                & \checkmark   && $\times$    &\\
Graph rotation            & \checkmark   &&&\\
Asymmetric error bars     & $\times$     &&&\\
Log axis                  & $\times$     &&&\\
Dates                     & $\times$     && \checkmark &\\
Entry of metadata         & \checkmark   &&&\\
Calculation of mean/sd/n  & \checkmark   & $\times$    &            & \\
Automated point detection & $\times$     &            &            & \\
\hline
\end{tabular}

\end{table}

@@

\subsection{Scatter plots}
\fct{metaDigitise} prompts the user to enter groups names and then to click on points. Points added by mistake can be deleted. The user can subsequently add groups, edit groups (add or remove points) or delete groups. Different groups are plotted in different colours and shapes, with a legend at the bottom of the figure (Figure \ref{fig:all_extract}C). Mean, SD and sample size (n) are calculated from the clicked points, for each group. In the case that the sample size from the clicked points does not match a known sample size (e.g. if there are overlaid points), the user can enter a alternate sample size.

\subsection{Histograms}
\fct{metaDigitise} prompts the user to click on the top corners of each bar. Bars can subsequently be deleted. 
For each bar a midpoint (mean x coordinates) and a frequency (mean y coordinates, rounded to the nearest integer) is calculated. The sample size is calculated as the sum of the frequencies., the mean as
\begin{equation}
\sigma = SE * \sqrt{n}
\end{equation}
and from 95\% confidence intervals as
\begin{equation}
\sigma = \frac{CI}{1.96} * \sqrt{n}
\end{equation}

Using the x coordinates, \fct{metaDigitise} then calculates the midpoint and the frequency for each bar.

@@

We'll now demonstrate how \fct{metaDigitise} works using figures generated from the well known iris data set. Users can install the \pkg{metaDigitise} package from GitHub as follows:

@@

\section{Conclusions and Future Enhancements}

\pkg{metaDigitise} is available through GitHub. WHile it is incredibly flexible already, our future plans include building in abilities to deal with log-transformed axes, including arguments for calculating standard deviations from 95\% confidence intervals and standard errors using the t-distribution to correct the t-value for small sample sizes (currently assumed t-values are 1.96 as is normal), dealing with asymmetric error bars and the possibility of zooming in plots such that greater accuracy can be achieved when digitising. We hope to also provide options for assessing inter-observer reliability in the future.

% maybe something about automated point detection - from our experience, as calibration needs to be done for each plot, and often points are not detected, it is much more reliable, and sometimes even quicker to manuslly select the points

>>>>

%% recommended packages
\usepackage{thumbpdf,lmodern,amssymb,rotating}

% \usepackage[utf8]{inputenc},
% \usepackage{geometry} 
% \geometry{a4paper}  
% %\usepackage[parfill]{parskip} 
% \usepackage{graphicx}	
% \usepackage{amsmath}
% \usepackage{fullpage}
% \usepackage{setspace} 
% \usepackage{lineno}
% \usepackage[none]{hyphenat}
% \usepackage{listings}
% \usepackage[round]{natbib}

@@

In many different contexts, researchers need to make use of data presented in primary literature. Most notably, this includes meta-analysis, which is becoming increasingly common in many research fields. Meta-analysis uses effect size estimates and their sampling variance, taken from many studies, to understand whether particular effects are common across studies and to explain variation among these effects \citep{Naka2017, Glass1976, Koricheva2013, Borenstein2009}. Meta-analysis therefore relies foremost on data extracted from primary literature and more specifically statistics that have been reported in the text or tables of research papers. These summary statistics are also, however, frequently presented in figures and so need to be manually extracted using digitising programs. Although there are several existing tools to perform tasks like this (e.g. DataThief \citep{DataThief}, GraphClick \citep{GraphClick}, WebPlotDigitizer \citep{WebPlotDigitizer}), these tools are not designed this specific purpose (i.e. a meta-analysis). 
%% need to spell out clear points
%% 1 - dont manipulate data/allow different plot types
Firstly, they typically only provide the user with calibrated x,y coordinates from imported figures, and do not differentiate between common plot types that are use to present data. This means that a large amount of downstream data manipulation is subsequently required. For example, data are frequently presented in mean and error plots (Figure \ref{fig:plot_type}A), for which the user wants a mean and error estimate for each group presented in the figure. With existing programs, x,y coordinates of means and errors are returned, to which the user must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to a standard deviation, depending on the type of error presented.
%% 2- adding metadata
Secondly, digitising programs do not easily allow the integration of metadata at the time of data extraction, such as  grouping points or adding variable names, group names and sample sizes.
%% 3 - not in statistical software
Additionally, existing stand-alone software does not allow easy import into commonly used statistical software (such as \proglang{R}), or try to optimize the research pipelines to facilitate editing and reproducibility. These are major issues because extracting from figures can be an incredibly time-consuming process. Furthermore, although meta-analysis is an important tool in consolidating the data from many studies, relying on transparency of those studies, many of the processes involved in data extraction are opaque and difficult to reproduce, making extending studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and go a long way to resolving the reproducibility crises we are seeing in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, sandve_ten_2013}.

@@

@@

\subsection{Scatter plots}
\fct{metaDigitise} prompts the user to enter groups names and then to click on points. Points added by mistake can be deleted. The user can subsequently add groups, edit groups (add or remove points) or delete groups. Different groups are plotted in different colours and shapes, with a legend at the bottom of the figure (Figure \ref{fig:all_extract}C). Mean, SD and sample size (n) are calculated from the clicked points, for each group. In the case that the sample size from the clicked points does not match a known sample size (e.g. if there are overlaid points), the user can enter an alternate sample size.

\subsection{Histograms}
\fct{metaDigitise} prompts the user to click on the top corners of each bar. Bars can subsequently be deleted. 
For each bar a midpoint (m; mean x coordinates) and a frequency (f; mean y coordinates, rounded to the nearest integer) is calculated. The sample size, mean and SD were calculated as:
\begin{equation}
n = \sum_{i=1}^n{f_i}
\end{equation}

\begin{equation}
\mu = \frac{\sum_{i=1}^n{m_i f_i}}{n}
\end{equation}

\begin{equation}
\sigma = \sqrt{\frac{\sum_{i=1}^n{(m_i f_i - \mu f_i)^2}}{n-1}}
\end{equation}

As with the scatterplots, If the sample size from the extracted data does not match a known sample size, the user can enter an alternate sample size.

@@

We will now demonstrate how \fct{metaDigitise} works using figures generated from the well known iris data set. Users can install the \pkg{metaDigitise} package from GitHub as follows:

@@

\section{Discussion and Conclusions}

%% discuss specifically everything it can't do, and jsutify it slightly

\begin{sidewaystable}[t!]
\centering
\begin{tabular}{lccccccc}
\hline
Function                  & metaDigitise & GraphClick$^1$ & DataThief$^2$  & DigitizeIt$^3$ & WebPlotDigitizer$^4$ & metagear$^5$ & digitize$^6$\\
\hline
Scatterplots              & \checkmark   & \checkmark & \checkmark & \checkmark & \checkmark     & \checkmark$^7$ & \checkmark \\
Mean and error plots      & \checkmark   & \checkmark & \checkmark & $\times$   & $\times$       & \checkmark$^7$ & $\times$    \\
Boxplots                  & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Histograms                & \checkmark   & $\times$   & $\times$   & $\times$   & \checkmark$^7$ & $\times$       & $\times$    \\
Graph rotation$^8$        & \checkmark   & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Groups                    & \checkmark   & \checkmark & $\times$   & \checkmark & \checkmark     & $\times$       & $\times$    \\
Entry of metadata         & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Summarising data          & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Multiple image processing & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Editing digitisation$^9$  & \checkmark   & \checkmark & \checkmark & $\times$   & \checkmark     & $\times$       & $\times$    \\
Automated point detection & $\times$     & \checkmark &     ?      & \checkmark & \checkmark     & \checkmark     & $\times$    \\
Line extraction           & $\times$     & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Zoom                      & $\times$     & \checkmark & \checkmark & \checkmark &       ?        & $\times$       & $\times$    \\
Log axis                  & $\times$ *   & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Dates                     & $\times$ *   & $\times$   & \checkmark & $\times$   & \checkmark     & $\times$       & $\times$    \\
Asymmetric error bars     & $\times$     & $\times$   & \checkmark & $\times$   & $\times$       & $\times$       & $\times$    \\
Freeware                  & \checkmark$^{10}$ & \checkmark$^{11}$ & \checkmark$^{11}$ & $\times$$^{11}$ & \checkmark$^{11}$ & \checkmark$^{10}$ & \checkmark$^{10}$\\
\hline

%http://datathief.org/DatathiefManual.pdf
%http://www.digitizeit.de/#Features
%https://automeris.io/WebPlotDigitizer/userManual.pdf

\end{tabular}

$^1$ \citet{GraphClick}
$^2$ \citet{DataThief}
$^3$ \citet{DigitizeIt}
$^4$ \citet{WebPlotDigitizer}
$^5$ \citet{Lajeunesse2016}
$^6$ \citet{Poisot2011}
$^7$ Only automated, no manual extraction.
$^8$ Or handles rotated graphs. 
$^9$ Allows saving, re-plotting and editing of data extraction.
$^{10}$ \proglang{R} package.
$^{11}$ Standalone software.

\end{sidewaystable}

%\pkg{metaDigitise} is available through GitHub. WHile it is incredibly flexible already, our future plans include building in abilities to deal with log-transformed axes, including arguments for calculating standard deviations from 95\% confidence intervals and standard errors using the t-distribution to correct the t-value for small sample sizes (currently assumed t-values are 1.96 as is normal), dealing with asymmetric error bars and the possibility of zooming in plots such that greater accuracy can be achieved when digitising. We hope to also provide options for assessing inter-observer reliability in the future.

Although \pkg{metaDigitise} is already very flexible, and provides functionality not seen in any other package, from comparison with other packages (Table \ref{tab:comparison}) it is clear that there are some functions that it does not perform. A notable feature that \pkg{metaDigitise} lacks is automated point detection. This is available in several packages (Table \ref{tab:comparison}). However, from our experience of using these functions, as calibration (for point detection) needs to be done for each plot individually, and often some points are not detected, it is much more reliable, and sometimes even quicker to manually select the points. From the testing of our package (see above) figures can be extracted in around 1-2 minutes, including entry of metadata. We do not think that current automation methods give substantial benefits to this.

Another feature that \pkg{metaDigitise} (currently) lacks, if a zoom function. This may enable users to gain greater accuracy when clicking on points. However, from our own experience, if you are using a reasonably sized screen then the accuracy is already high from these programs (see results above for example), and not much gain is to be had from zooming in on points.

\pkg{metaDigitise} also does not extract lines from figures, as many other packages do. In our own experience, this is not particularly useful for meta-analysis, although we recognise that it may be useful in other fields. In this case we would recommend extracting data as a scatter plot, and clicking along the line in question. A model can then be fitted to these points to estimate the parameters needed.

Finally, \pkg{metaDigitise} does not allow for asymmetric error bars. At present this is a deliberate omission, as it is not clear how best to derive SD from such data, given also that such asymmetric error bars may represent different things in different figures.
